Towards Practical First-Order Model Counting: Technical Appendix

### 2025-01-11 doc/old_paper/technical_appendix.tex
First draft of the SAT version

\documentclass{article}
\pdfpagewidth=8.5in
\pdfpageheight=11in

\usepackage{kr}

\usepackage{times}
\usepackage{soul}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[small]{caption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
\urlstyle{same}

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage[capitalize,noabbrev]{cleveref}
\usepackage{microtype}
\usepackage{mathtools}
\usepackage{forest}
\usepackage{siunitx}
\usepackage{amsfonts}
\usepackage{pifont}
\usepackage[inline]{enumitem}

\usetikzlibrary{arrows.meta}
\usetikzlibrary{positioning}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{fact}{Observation}
\newtheorem{corollary}{Corollary}
\newtheorem{assumption}{Assumption}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\pdfinfo{
/TemplateVersion (KR.2022.0, KR.2023.0, KR.2024.0)
}

\forestset{
  sn edges/.style={for tree={edge={-Latex}}}
}

\newcommand{\crefrangeconjunction}{--}

\crefname{line}{line}{lines}

\crefname{fact}{Observation}{Observations}
\crefname{assumption}{Assumption}{Assumptions}

\crefalias{enumi}{type}
\crefname{type}{Type}{Types}
\creflabelformat{type}{#2\textup{#1}#3}

\crefalias{enumi}{step}
\crefname{step}{Step}{Steps}
\creflabelformat{step}{#2\textup{#1}#3}

\crefalias{clause}{equation}
\crefname{clause}{Clause}{Clauses}
\creflabelformat{clause}{#2\textup{(#1)}#3}

\crefalias{formula}{equation}
\crefname{formula}{Formula}{Formulas}
\creflabelformat{formula}{#2\textup{(#1)}#3}

\DeclareMathOperator{\CR}{CR}
\DeclareMathOperator{\DR}{DR}
\DeclareMathOperator{\Reff}{Ref}

\DeclareMathOperator{\Doms}{Doms}

\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

\newcommand{\expr}{\mathtt{expr}}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\FO}{$\mathsf{FO}$}
\newcommand{\UFO}{$\mathsf{UFO}^{2} + \mathsf{EQ}$}

\newcommand{\Cranetwo}{\textsc{Crane2}}
\newcommand{\Cranebfs}{\textsc{Crane2-BFS}}
\newcommand{\Cranegreedy}{\textsc{Crane2-Greedy}}

\SetKwFunction{CompileWithBaseCases}{CompileWithBaseCases}
\SetKwFunction{Compile}{Compile}
\SetKwFunction{Propagate}{Propagate}
\SetKwFunction{FindBaseCases}{FindBaseCases}
\SetKwFunction{Simplify}{Simplify}

\title{Towards Practical First-Order Model Counting}

\author{%
  Ananth K. Kidambi$^1$\footnote{The first and second authors contributed equally and were affiliated with the National University of Singapore during the completion of this work.}\and
Guramrit Singh$^1$\and
Paulius Dilkas$^2$\and
Kuldeep S. Meel$^3$ \\
\affiliations
$^1$Indian Institute of Technology Bombay, Mumbai, India\\
$^2$National University of Singapore, Singapore, Singapore\\
$^3$University of Toronto, Toronto, Canada\\
\emails
210051002@iitb.ac.in,
guramrit@iitb.ac.in,
paulius.dilkas@nus.edu.sg,
meel@cs.toronto.edu
}% TODO: update affiliation and email

@@

\begin{abstract}
  First-order model counting (FOMC) is the problem of counting the number of
  models of a sentence in first-order logic. Recently, a new algorithm for FOMC
  was proposed that, instead of simply providing the final count, generates
  definitions of (possibly recursive) functions, which can be evaluated with
  different arguments to compute the model count for any domain size. However,
  the algorithm did not include base cases in the recursive definitions. This
  work makes three contributions. First, we demonstrate how to construct
  function definitions that include base cases by modifying the logical formulas
  used in the FOMC algorithm. Second, we extend the well-known circuit
  modification technique in knowledge compilation, known as smoothing, to work
  with the formulas corresponding to base cases and the recently proposed node
  types, such as domain recursion and constraint removal. Third, we introduce a
  compilation algorithm that transforms the function definitions into C++ code,
  equipped with arbitrary-precision arithmetic. These additions allow the new
  FOMC algorithm to scale to domain sizes over \num{9000} times larger than the
  current state of the art, as demonstrated through experimental results.
\end{abstract}

% TODO: No proper runtime consideration (other than the termination). (address
% this criticism in some way)

% TODO: Having an algorithm that can efficiently handle large domain sizes is
% crucial in the weighted setting. For example, consider the well-known "friends
% and smokers" Markov logic network [1]. The utility of such a model would be
% significantly limited if exact probabilities could only be efficiently
% computed for social networks of at most 30 people.
% [1] Richardson, M., & Domingos, P. (2006). Markov logic networks. Machine learning, 62, 107-136.

% TODO: somehow indicate that source code will be available (a redacted link?)

% TODO: Updates to the algorithm
% 1) replace x_i-c_i on Line 7 of Algorithm 2 with a generalisation of x_i that can be an arbitrary summation/subtraction of domain sizes
% 2) a function call with a constant should always mean that that function should have a base case, not just when it's in the same function

% TODO: must find a way to include the implementation in the submission

% TODO: consider restructuring the paper to make it easier to understand

\section{Introduction}
% TODO: improvements to the introduction (must):
% * make it extra clear what the gap is that this work is trying to fill
% * mention Crane earlier and describe its importance
% * the last two paragraphs should make more of a BOOM (e.g., include the over 9k claim)
% * (here and elsewhere) clarify the distinction between the weighted and unweighted variants of the problem in the final version of the paper. We focus on the unweighted variant in our implementation and experiments, however, Crane2 can handle WFOMC instances as well simply by switching the main data type from integers to rational numbers. Maybe add some weighted experiments as well.
% * (must) more related work, situating my contributions in the broader context
% * (concerning related work) the topic is intimately related to "graded logics", largely investigated in DL, mu-calculus and temporal logics

% 1. What is the problem?

\emph{First-order model counting} is the task of counting the number of models
of a sentence in first-order logic over some given domain(s). The (symmetric)
weighted variant of this problem, known as WFOMC, seeks to calculate the sum of
model weights. In WFOMC, the weight of a model is determined by predicate
weights~\cite{DBLP:conf/ijcai/BroeckTMDR11}. WFOMC is a key approach to
\emph{lifted (probabilistic) inference}, which aims to compute probabilities
more efficiently by leveraging symmetries inherent in the problem
description~\cite{DBLP:conf/ecai/Kersting12}.

% 2. Why is it interesting and important?

Lifted inference is an active area of research, with recent work in domains such
as constraint satisfaction problems~\cite{DBLP:journals/jair/TotisDRK23} and
probabilistic answer set programming~\cite{DBLP:journals/ijar/AzzoliniR23}.
WFOMC has been used for inference on probabilistic
databases~\cite{DBLP:journals/debu/GribkoffSB14} and probabilistic logic
programs~\cite{DBLP:journals/ijar/RiguzziBZCL17}. By considering domains of
increasing sizes, the model count of a formula can be seen as an integer
sequence. WFOMC algorithms have been utilised for discovering new
sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23} as well as
conjecturing~\cite{DBLP:conf/ilp/BarvinekB0ZK21} and
constructing~\cite{DBLP:conf/kr/DilkasB23} recurrence relations and other
recursive structures that describe these sequences. Additionally, WFOMC
algorithms have been extended to perform
\emph{sampling}~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}.

% 3. Why is it hard? (E.g., why do naive approaches fail?)

The complexity of WFOMC is typically characterised in terms of \emph{data
  complexity}. This involves fixing the formula and determining whether an
algorithm exists that can compute the WFOMC in time polynomial with respect to
the domain size(s). If such an algorithm exists, the formula is called
\emph{liftable}~\cite{DBLP:conf/starai/JaegerB12}. Beame et
al.~\shortcite{DBLP:conf/pods/BeameBGS15} demonstrated the existence of an
unliftable formula with three variables. It is also known that all formulas with
up to two variables are
liftable~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}. The liftable
fragment of formulas with two variables has been expanded with various
axioms~\cite{DBLP:conf/aaai/TothK23,DBLP:journals/ai/BremenK23}, counting
quantifiers~\cite{DBLP:journals/jair/Kuzelka21} and in other
ways~\cite{DBLP:conf/nips/KazemiKBP16}.

% 4. Why hasn't it been solved before? (Or, what's wrong with previous proposed
% solutions? How does mine differ?)

There are various WFOMC algorithms with different underlying principles. Perhaps
the most prominent class of WFOMC algorithms is based on \emph{first-order
  knowledge compilation}. In this approach, the formula is compiled into a
representation (such as a circuit or graph) by iteratively applying
\emph{compilation rules}. This representation can then be used to compute the
WFOMC for any combination of domain sizes (or weights). Algorithms in this class
include \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its recent
extension \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. The former compiles
formulas into circuits, while the latter compiles them first to
\emph{first-order computational graphs} (FCGs) and then to (algebraic)
equations. Another WFOMC algorithm,
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}, is based on cell enumeration.
Other algorithms utilise local search~\cite{DBLP:journals/pvldb/NiuRDS11},
junction trees~\cite{DBLP:conf/aaai/VenugopalSG15}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime approximation via
upper/lower bound construction~\cite{DBLP:conf/ijcai/BremenK20}.
% TODO: FCGs were never defined. Maybe avoid referring to them altogether?
% Smoothing nodes are also undefined.

% 5. What are the key components of my approach and results? Also include any
% specific limitations.

% TODO: first sentence: elaborate a bit more for a general audience
The recently proposed \textsc{Crane} algorithm, while capable of handling
formulas beyond the reach of \textsc{ForcLift}, was incomplete as it could only
construct function definitions, which would then need to be evaluated to compute
the WFOMC.\@ Additionally, recursive functions were presented without base
cases. In this work, we introduce \Cranetwo{}, an extension of \textsc{Crane}
that addresses these two weaknesses.
% TODO: We called the previous version of Crane "incomplete" because it cannot
% be used as a black box that takes Markov logic networks (or weighted formulas
% in some other format) and outputs numbers (i.e., model counts or
% probabilities).

\begin{figure*}[t]
  \centering
  \begin{tikzpicture}
    \node at (-1, 0) (formula) {$\phi$};
    \node[draw,rounded rectangle] at (3, 0) (compilewithbasecases) {\CompileWithBaseCases};
    \node[draw,rounded rectangle] at (9, 0) (compilation) {Compile to C++};

    \node[draw,rounded rectangle,dashed] at (12, 0) (cpp) {C++ code};
    \node at (12, -1) (sizes) {Domain sizes};

    \node at (15, 0) (count) {Model count};

    \node[draw,rounded rectangle] at (3, -2) (findbasecases) {\FindBaseCases};
    \node[draw,rounded rectangle,left = 0.1cm of findbasecases] (crane) {\Compile};
    \node[draw,rounded rectangle,right = 0.1cm of findbasecases] (propagate) {\Propagate};
    \node[draw,rounded rectangle,right = 0.1cm of propagate] (simplify) {\Simplify};

    \node[draw,fit={(compilewithbasecases) (compilation) (crane) (findbasecases) (propagate)},inner ysep=7pt,yshift=5pt] {};
    \node at (0.2, 0.5) {\Cranetwo};

    \draw[-Latex] (formula) -- (compilewithbasecases);
    \draw[-Latex] (compilewithbasecases) -- node[above] {$\mathcal{E}$} (compilation);
    \draw[-Latex] (compilation) -- (cpp);
    \draw[-Latex] (sizes) -- (cpp);
    \draw[-Latex] (cpp) -- (count);

    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (findbasecases);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (propagate);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,right] {uses} (simplify);
  \end{tikzpicture}
  \caption[]{The outline of using \Cranetwo{} to compute the model count of a formula $\phi$. First, the formula is compiled into a set of equations, which are then used to create a C++ program. This program can be executed with different command line arguments to calculate the model count of $\phi$ for different domain sizes. To accomplish this, the \CompileWithBaseCases function employs several components:
    \begin{enumerate*}[label=(\roman*)]
      \item the knowledge compilation algorithm of \textsc{Crane}, referred to as \Compile,
      \item a procedure called \FindBaseCases, which identifies a sufficient set of base cases,
      \item a procedure called \Propagate, which constructs a formula corresponding to a given base case, and
      \item algebraic simplification techniques (denoted as \Simplify).
    \end{enumerate*}
  }\label{fig:overview}
\end{figure*}

\Cref{fig:overview} outlines the workflow of the new algorithm. In
\cref{sec:main}, we describe how \CompileWithBaseCases finds the base cases for
recursive functions by:
\begin{enumerate*}[label=(\roman*)]
  \item identifying a sufficient set of base cases for each function,
  \item constructing formulas corresponding to these base cases,
  and\label[step]{step2}
  \item recursing on these new formulas.
\end{enumerate*}
Then, \cref{sec:smoothing} explains post-processing techniques for FCGs and the
formulas from \cref{step2} above required to preserve the correct model count.
Next, \cref{sec:cpp} elucidates how function definitions encoding a solution to
a WFOMC problem are compiled into C++ programs.\footnote{The possibility of
  compiling WFOMC solutions to C++ programs has been previously
  considered~\cite{DBLP:conf/kr/KazemiP16}. However, the extent of formulas that
  could be handled was limited.} Finally, \cref{sec:experiments} presents two
experiments comparing \Cranetwo{} with other state-of-the-art WFOMC algorithms.

\section{Preliminaries}

In \cref{sec:logic}, we provide a summary of the basic principles of first-order
logic. Then, in \cref{sec:threelogics}, we formally define WFOMC and discuss the
distinctions between three variations of first-order logic that are utilised for
WFOMC.\@ Finally, in \cref{sec:algebra}, we introduce the terminology used to
describe the output of the original \textsc{Crane} algorithm, i.e., functions
and equations that define them.

We use $\mathbb{N}_{0}$ to represent the set of non-negative integers. In both
algebra and logic, we write $S\sigma$ to denote the application of a
\emph{substitution} $\sigma$ to an expression $S$, where
$\sigma = [x_{1} \mapsto y_{1}, x_{2} \mapsto y_{2}, \dots, x_{n} \mapsto y_{n}]$
signifies the replacement of all instances of $x_{i}$ with $y_{i}$ for all
$i = 1, \dots, n$.

\subsection{First-Order Logic}\label{sec:logic}

In this section, we will review the basic concepts of first-order logic as they
are used in first-order knowledge compilation algorithms. There are two key
differences between the logic used within such algorithms and the logic
supported as input. First, Skolemization~\cite{DBLP:conf/kr/BroeckMD14} is
employed to eliminate existential quantifiers by introducing additional
predicates. Second, the input formula is rewritten as a conjunction of clauses,
each of which is in \emph{prenex normal form}~\cite{hinman2018fundamentals}.
% TODO: explain that "Skolemization" here is NOT standard Skolemization of
% first-order logic, which would replace existential quantifiers by function
% terms.

% TODO: reorder the definitions in the opposite direction
A \emph{formula} is a conjunction of clauses. A \emph{clause} is of the form
$\forall x_{1} \in \Delta_{1}\text{.}\forall x_{2} \in \Delta_{2}\dots\text{
}\forall x_{n} \in \Delta_{n}\text{.}\phi(x_{1}, x_{2}, \dots, x_{n})$, where
$\phi$ is a disjunction of literals that only contain variables
$x_{1}, \dots, x_{n}$ (and any constants). We say that a clause is a
\emph{(positive) unit clause} if
\begin{enumerate*}[label=(\roman*)]
  \item there is only one literal with a predicate, and
  \item it is a positive literal.
\end{enumerate*}
A \emph{literal} is either an atom (i.e., a \emph{positive} literal) or its
negation (i.e., a \emph{negative} literal). An \emph{atom} is either
\begin{enumerate*}[label=(\roman*)]
  \item $P(t_{1}, \dots, t_{m})$ for some predicate $P$ and terms
  $t_{1}, \dots, t_{m}$ (written as $P(\mathbf{t})$ for short) or
  \item $x=y$ for some terms $x$ and $y$.
\end{enumerate*}
An atom is \emph{ground} if it contains no variables, i.e., only constants. The
\emph{arity} of a predicate is the number of arguments it takes, i.e., $m$ in
the case of predicate $P$ mentioned above. When we want to denote a predicate
together with its arity, we write $P/m$. A \emph{term} is either a variable or a
constant. Throughout the paper, we will use set-theoretic notation, interpreting
a formula as a set of clauses and a clause as a set of literals. Moreover, for
readability, clauses written on separate lines are implicitly conjoined.

\subsection{WFOMC Algorithms and Their Logics}\label{sec:threelogics}

% TODO: mention the Unique Name Assumption and that, hence, s = t holds iff s
% and t are the same constant.

\begin{table*}[t]
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{Comparison of three logics used in WFOMC based on the following aspects:
    \begin{enumerate*}[label=(\roman*)]
      \item number of sorts,
      \item support for constants,
      \item maximum number of variables,
      \item supported quantifiers, and
      \item supported atoms in addition to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and $n$-tuple of terms $\mathbf{t}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k$ and $m$ are non-negative integers, with the latter depending on the domain size,
      \item $P$ represents a predicate, and
      \item $x$ and $y$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

% TODO:
% 1. Rename \UFO{} to what that new paper uses
% 2. Use FastWFOMC.jl instead, removing the mention of 'privately-obtained'
% 3. use symbols instead of numbers for the footnotes
In \cref{tbl:logics}, we outline the differences among three first-order logics
commonly used in WFOMC:
\begin{enumerate*}[label=(\roman*)]
  \item \FO{} is the input format for
  \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}} and
  its extensions
  \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}} and
  \Cranetwo{};
  \item \Ctwo{} is often used in the literature on \textsc{FastWFOMC} and
  related
  methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22};
  \item \UFO{} is the input format supported by a private version of
  \textsc{FastWFOMC} obtained directly from the authors.
\end{enumerate*}
Note that the publicly available
version\footnote{\url{https://comp.nus.edu.sg/~tvanbr/software/fastwfomc.tar.gz}}
of \textsc{FastWFOMC} does not support any cardinality constraints. The notation
we use to refer to each logic is standard in the case of \Ctwo{}, new in the
case of \UFO{}, and redefined to be more specific in the case of \FO{}. All
three logics are function-free, and domains are always assumed to be finite.

In \FO{}, each term is assigned to a \emph{sort}, and each predicate $P/n$ is
assigned to a sequence of $n$ sorts. Each sort has its corresponding domain.
Most of these assignments to sorts are typically left implicit and can be
reconstructed from the quantifiers. For example, $\forall x,y \in \Delta\text{.
}P(x, y)$ implies that variables $x$ and $y$ have the same sort. On the other
hand, $\forall x \in \Delta\text{. }\forall y \in \Gamma\text{. } P(x, y)$
implies that $x$ and $y$ have different sorts, and it would be improper to
write, for example, $\forall x \in \Delta\text{. }\forall y \in \Gamma\text{.
} P(x, y) \lor x = y$. \FO{} is also the only logic to support constants,
formulas with more than two variables, and the equality predicate.
% TODO: the discussion about sorts isn't featured in the rest of the paper

\begin{remark}
  In the case of \textsc{ForcLift} and its extensions, support for a formula as
  valid input does not imply that the algorithm can compile the formula into a
  circuit or graph suitable for lifted model counting. However, it is known that
  \textsc{ForcLift} compilation is guaranteed to succeed on any \FO{} formula
  without constants and with at most two
  variables~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}.
\end{remark}

\Ctwo{} and \UFO{} are single-sorted (i.e., all variables are quantified over
the same domain) and only include formulas with at most two variables and no
constants. Unlike \UFO{}, \Ctwo{} supports existential quantifiers, including
\emph{counting quantifiers}. For example, $\exists^{=1} x\text{. }\phi(x)$ means
that there exists \emph{exactly one} $x$ such that $\phi(x)$, and
$\exists^{\le 2} x\text{. }\phi(x)$ means that there exist \emph{at most two}
such $x$. The advantage \UFO{} brings over \Ctwo{} is the support for
\emph{(equality) cardinality constraints}. For example, $|P| = 3$ constrains all
models to have \emph{precisely three positive literals with the predicate $P$}.

% TODO: The distinctness of domains is important in two ways. First, in terms of
% expressiveness, writing clause such as "\forall x, P(x, x)" is valid if
% predicate P is defined over two copies of the same domain and invalid
% otherwise. Second, having more distinct domains makes the problem more
% decomposable for the knowledge compilation algorithm. With more distinct
% domains, it can make assumptions or deductions about, e.g., the first domain
% of predicate P without worrying how (or if) they apply to the second domain.
\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For each predicate $P/n$ in $\phi$, let
  ${(\Delta_{i}^{P})}_{i=1}^{n}$ be a list of the corresponding domains (which
  may not be distinct). Let $\sigma$ be a map from the domains of $\phi$ to
  their interpretations as sets, satisfying the following conditions:
  \begin{enumerate*}[label=(\roman*)]
    \item the sets are pairwise disjoint, and
    \item the constants in $\phi$ are included in the corresponding domains.
  \end{enumerate*}
  Then a \emph{structure} of $\phi$ (with respect to $\sigma$) is a set $M$ of
  ground literals defined by adding either $P(\mathbf{t})$ or
  $\neg P(\mathbf{t})$ for every predicate $P/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{P})$. A structure is a
  \emph{model} if it satisfies $\phi$.
\end{definition}
% TODO: adding to what

% TODO: “Weights in Q” undefined, C2 undefined (and also far from well explained
% before), endofunction undefined (also endomorphisms missing), weight function
% undefined, weighted model count imprecise (above it’s called weighted
% first-order model count/wfmoc),
\begin{definition}[WFOMC~\cite{DBLP:conf/ijcai/BroeckTMDR11}]
  Continuing from \cref{def:model}, for each predicate $P$ in $\phi$, let
  $w^{+}(P)$ and $w^{-}(P)$ be its \emph{weights} in $\mathbb{Q}$. Unless
  specified otherwise, we assume all weights are equal to 1. The
  \emph{(symmetric) weighted first-order model count} (WFOMC) of $\phi$ (with
  respect to $\sigma$, $w^{+}$, and $w^{-}$) is:
  \[
    \sum_{M \models \phi} \prod_{P(\mathbf{t}) \in M} w^{+}(P) \prod_{\neg P(\mathbf{t}) \in M} w^{-}(P),
  \]
  where the sum is over all models of $\phi$.
\end{definition}

\begin{example}[Counting functions]\label{example:functions}
  To define predicate $P$ as an endofunction on $\Delta$, in \Ctwo{} one would
  write $\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta\text{.
  }P(x, y)$. In \UFO{}, the same could be written as
  \begin{equation}\label[formula]{eq:functions1}
    \begin{gathered}
      \forall x, y \in \Delta\text{. }S(x) \lor \neg P(x, y)\\
      |P| = |\Delta|,
    \end{gathered}
  \end{equation}
  where $w^{-}(S) = -1$. Although \cref{eq:functions1} has more models compared
  to its counterpart in \Ctwo{}, the weight function makes the \emph{weighted}
  model count of \cref{eq:functions1} equal to the number of endofunctions on
  $\Delta$.

  % TODO: Yes, as mentioned at the end of Section 2.1, clauses c and d written
  % on separate lines are to be interpreted as the conjunction of c and d (maybe
  % move that sentence here?).
  Equivalently, in \FO{} we would write
  \begin{equation}\label[formula]{eq:fo}
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z.
    \end{gathered}
  \end{equation}
  The first clause asserts that each $x$ must have at least one corresponding
  $y$, while the second statement adds the condition that if $x$ is mapped to
  both $y$ and $z$, then $y$ must equal $z$. It is important to note that
  \cref{eq:fo} is written with two domains instead of just one. However, we can
  still determine the number of endofunctions by assuming that the sizes of
  $\Gamma$ and $\Delta$ are equal. This formulation, as observed by Dilkas and
  Belle~\shortcite{DBLP:conf/kr/DilkasB23}, can prove beneficial in enabling
  first-order knowledge compilation algorithms to find efficient solutions.
\end{example}

\subsection{Algebra}\label{sec:algebra}

We write $\expr{}$ to represent an arbitrary algebraic expression. It is
important to note that some terms have different meanings in algebra compared to
logic. In algebra, a \emph{constant} refers to a non-negative integer. Likewise,
a \emph{variable} can either be a parameter of a function or a variable
introduced through summation, such as $i$ in the expression
$\sum_{i=1}^{n} \expr$. A (function) \emph{signature} is
$f(x_{1}, \dots, x_{n})$ (or $f(\mathbf{x})$ for short), where $f$ represents an
$n$-ary function, and each $x_{i}$ represents a variable. An \emph{equation} is
$f(\mathbf{x}) = \expr{}$, with $f(\mathbf{x})$ representing a signature.

\begin{definition}[Base case]\label{def:basecase}
  Let $f(\mathbf{x})$ be a function call where each $x_{i}$ is either a constant
  or a variable (note that signatures are included in this definition). Then
  function call $f(\mathbf{y})$ is considered a \emph{base case} of
  $f(\mathbf{x})$ if $f(\mathbf{y}) = f(\mathbf{x})\sigma$, where $\sigma$ is a
  substitution that replaces one or more $x_{i}$ with a constant.
\end{definition}

\section{Completing the Definitions of Recursive Functions}\label{sec:main}
% TODO: As mentioned in the caption of Figure 1, "Compile" refers to the
% knowledge compilation algorithm of Dilkas and Belle (2023). Its three return
% values are explained at the beginning of page 4. We do not elaborate on the
% details of "Simplify" as it primarily consists of elementary algebraic
% simplifications such as "1x = x" and "0x = 0". However, we highlight one
% non-trivial aspect of "Simplify" right before Section 3.1. We will improve the
% structure of Section 3 to make this information more readily available. Maybe
% write Crane instead of Compile?

% TODO: It would not be entirely accurate to call the algorithms presented in
% this work a "post-processing step" because Algorithm 1 is recursive. In other
% words, this "post-processing step" to first-order knowledge compilation can
% lead to more compilation that then updates/completes the original set of
% equations.
\begin{algorithm}[t]
  \caption{\protect\CompileWithBaseCases{$\phi$}}\label{alg:compilewithbasecases}
  \KwIn{formula $\phi$}
  \KwOut{set $\mathcal{E}$ of equations}
  $(\mathcal{E}, \mathcal{F}, \mathcal{D}) \gets \Compile{$\phi$}$\;\label{line:first}
  $\mathcal{E} \gets \Simplify{$\mathcal{E}$}$\;\label{line:second}
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$\mathcal{E}$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \ForEach{$i$ such that $x_{i} \in \mathbb{N}_{0}$}{
      $\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$\;
    }
    $\mathcal{E} \gets \mathcal{E} \cup \CompileWithBaseCases{$\psi$}$\;
  }
\end{algorithm}
% TODO: >> Alg 1, line 5: Why is this not an infinite loop? On line 5 of
% Algorithm 1, the loop iterates over the indices of \mathbf{x}. The arity of a
% function is always a finite number, typically equal to 1 or 2.

\Cref{alg:compilewithbasecases} presents our overall approach for compiling a
formula into equations that include the necessary base cases. To begin, we
utilise the knowledge compilation algorithm from the original \textsc{Crane} to
compile the formula into three components:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
After some algebraic simplification, $\mathcal{E}$ is then passed to the
\FindBaseCases procedure (explained in \cref{sec:identifying}), which identifies
the base cases that need to be defined. For each base case $f(\mathbf{x})$, we
determine the formula associated with the function name $f$ and simplify it
using the \Propagate procedure (explained in \cref{sec:simplifying}). The
algorithm then calls itself on these simplified formulas and adds the resulting
base case equations to $\mathcal{E}$. \Cref{example:overall} provides a more
detailed explanation of \cref{alg:compilewithbasecases}.

% TODO: F missing (also not explained that it’s not needed below). E is not explained other than that we take the example from X. Would not be a problem if it is motivated differently and the focus is on post-processing (defining inputs properly) and not on the entire algorithm.
% TODO: include a high-level overview of first-order knowledge compilation and the meaning of the resulting functions. For the most part, the compilation process that constructs these function definitions can be treated as a black box for the purposes of this work. However, a few high-level details still ought to be mentioned:
% 1. Each algebraic variable ("m", "n", and "l" in the example) corresponds to a domain size.
% 2. \mathcal{D} indicates that "m" denotes the size of \Gamma, "n" denotes the size of \Delta, etc.
% 3. \mathcal{E} can contain any (positive) number of functions.
% 4. Function denoted by "f" represents the solution to the model counting problem.
% 3. The model count of the original formula is computed by evaluating f on the desired domain sizes. For example, f(3, 2) is the model count when |\Gamma| = 3, and |\Delta| = 2.
\begin{example}[Counting bijections]\label{example:overall}
  Consider the following formula (previously examined by Dilkas and
  Belle~\shortcite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ as a
  bijection between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z\\
      \forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z.
    \end{gathered}
  \]
  We specifically examine the first solution returned by \Cranetwo{} for this
  formula.

>>>>

%File: anonymous-submission-latex-2025.tex
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage[submission]{aaai25}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet}  % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage[capitalize,noabbrev]{cleveref}

% Keep the \pdfinfo as shown here. There's no need
% for you to add the /Title and /Author tags.
\pdfinfo{
/TemplateVersion (2025.1)
}

\setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.

\SetKwFunction{CompileWithBaseCases}{CompileWithBaseCases}
\SetKwFunction{FindBaseCases}{FindBaseCases}

\newcommand{\Cranetwo}{\textsc{Crane2}}
\newcommand{\expr}{\mathtt{expr}}

\newtheorem{theorem}{Theorem}
\newtheorem{assumption}{Assumption}
\newtheorem{corollary}{Corollary}
\newtheorem{fact}{Observation}
\newtheorem{lemma}{Lemma}

\title{Towards Practical First-Order Model Counting: Technical Appendix}
\author{Anonymous Submission}
\affiliations{}

@@

\section{The Proof of Theorem 1}

To demonstrate that the base cases identified by \FindBaseCases are sufficient,
we begin with a few observations that stem from
previous~\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:conf/kr/DilkasB23} and this
work. Let $\mathcal{E}$ represent the equations returned by
\CompileWithBaseCases.
